Variance reduction for MCMC methods via martingale representations

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_L)-\psi(X_L)|]^2

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$. At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}. Thus, the ratio of cost-variances is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma=\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}$  

@@

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
\begin{eqnarray*}
A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
\end{eqnarray*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$. 
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0.
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias. Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}.
\end{equation}
Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.

>>>>

In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$. At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma<\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  

@@

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
\begin{eqnarray*}
A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
\end{eqnarray*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$. 
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0.
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias. Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias_2}
\tvnorm{\pi - \pi_\gamma} < C\sqrt{\gamma}.
\end{equation}
Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
